Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2024-10-04 paper.tex
Fix typo
Minor edits and clarifications

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear wave theory and derive higher-order
steady solutions for the change of short-wave wavenumber, action density, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
We further examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs only for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The combined analytical and numerical framework presented here allows
quantifying the importance of various terms in the full wave conservation
equations.
The proposed analytical solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx 0.2$.
As a corollary, except for a subset of initial conditions, the solutions to
fully nonlinear crest-action conservation equations are mostly steady in the
reference frame of the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to (\ref{eq:wavenumber}), except for the last term
which represents the inhomogeneity of the group speed.
As \citet{peureux2021unsteady} explained, the presence of this term causes
unstable growth of wave action in infinite long-wave trains.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are the governing equations
that we approximate and solve analytically in \S\ref{section:analytical_solutions},
and numerically in their full form in \S\ref{section:numerical_solutions}.
It is worth noting that this system of equations is semi-coupled, meaning that
it is possible to solve (\ref{eq:wavenumber}) on its own, but solving
(\ref{eq:wave_action2}) requires also the solution of (\ref{eq:wavenumber}).
In other words, the wave kinematics (the wavenumber distribution) are not
concerned with the wave action.
In contrast, the wave action is governed, among others, by the
group velocity and thus depends on the wavenumber.

@@

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\widetilde{g} = \left( 1 - \varepsilon_L \cos{\psi} \right)
\end{equation}
which is the form used by \citet{peureux2021unsteady}, for example.

@@

\subsection{Unstable growth in Pereux et al. (2021)}
\label{subsection:unstable_growth}

@@

Why is it, then, that the solution for short-wave modulation is not only
quantitatively, but also qualitatively, different depending on the short-wave
initial conditions?
It may at first seem that there is something special about the uniform initial
conditions that makes its solutions grow unsteadily and indefinitely.
\citet{peureux2021unsteady} described this scenario as "...the sudden appearance
of a long wave perturbation in the middle of a homogeneous short wave field".
However, such a scenario is not realistic in the ocean where waves tend to
organise into groups due to dispersion.
To test whether the sudden onset of long-wave forcing causes the unstable growth,
we repeat the numerical simulation of Fig.
\ref{fig:modulation_3panel_infinite}, except for a linear ramp applied to $a_L$
(and consequently, the other long-wave properties including the orbital
velocities and the local steepness) for the first five periods:
$a_L(t) = min(a_L, a_L t / \left(5 T_L\right))$, where $T_L$ is the long-wave period.
This result is shown in Fig. \ref{fig:modulation_3panel_ramp}.
The key distinction that arises with the introduction of the ramp is that the
modulation pattern of short waves is now locked-in to the long-wave crests,
and their indefinite steepening vanishes.
The stabilisation effect of the long-wave ramp is more readily assessed in
Fig. \ref{fig:unsteady_growth_timeseries}, which shows the maximum short-wave
steepness modulation as function of time over 30 long-wave periods, for the
infinite long-wave train case and the case of applying a linear ramp.
Comparing the wave action propagation ($C_g \frac{\partial N}{\partial x}$) and
inhomogeneity ($N \frac{\partial C_g}{\partial x}$) tendencies reveals that
after only one long-wave period, the propagation and inhomogeneity tendencies
are unbalanced and amplify the steepening of short waves
(Fig. \ref{fig:inhomogeneity_tendencies}).
\citet{peureux2021unsteady} correctly pointed out that the unstable
growth of short-wave steepness occurs due to the resonance of the inhomogeneity
tendency with the short-wave action perturbation.
However, in the case of a gentle ramp-up of the long-wave forcing, the
propagation and inhomogeneity tendencies mostly balance each other and allow
the short-wave steepening to lock-in to the long-wave crests.

@@

We now proceed to examine the numerical solutions with the long-wave forcing
applied as a long-wave group.
The wave group is simulated as a non-dispersive train of long waves whose
amplitude scales with $\sin[\pi t / (n T_L)]$, where $n$ is the number of
long waves in the group.
The elevation of the long waves thus gradually ramps up from rest to $a_L$ at
$t = N/2 T_L$, after which it gradually decreases toward zero.
In this simulation, $n = 10$ and $\varepsilon_L = 0.1$ as in the previous two
simulations.
The modulations of short-wave wavenumber, amplitude, and steepness in response
to such long-wave group is shown in Fig. \ref{fig:modulation_3panel_groups}.
Like in the case of a linear ramp, the modulation pattern here is locked in to
the long-wave crest and peaks at approximately 1.2 for short-wave steepness.
After the peak of the long-wave group passes, the short-wave modulation recedes
back toward the resting state.

@@

The solutions presented here, analytical and numerical alike, are valid only
for certain scales of short and long waves.
Specifically, the wave action conservation equation, introduced by
\citet{bretherton1968wavetrains}, requires that the ambient velocity is slowly
varying on the time scales of short waves, \textit{i.e.}
$|\partial U / \partial x| \ll k |U|$
and $|\partial U / \partial t| \ll \sigma |U|$.
This implies $k_L \ll k$ and $|\sigma_L| \ll \sigma|$, respectively.
Another requirement is for the short-waves to remain linear, which implies
$a k \ll 1$.
Furthermore, the short waves examined here do not include the effects of
surface tension (gravity-capillary waves) or intermediate or shallow water
(although the solutions can be extended to those conditions, as
\citet{phillips1981dispersion} did, for example).
Within these limits, the crest-action conservation equations provide an
elegant and relatively simple framework to study the hydrodynamic modulation
of short waves by longer waves.
This approach, of course, isolates this particular process from other
dominant modulation and nonlinear transfer processes, such as wave growth
by wind, wave dissipation, and aerodynamic sheltering by the longer waves,
that are otherwise present in most measurements
\citep{plant1986two,laxague2017gravity}.

@@

Despite several studies that provided qualitatively similar steady solutions
for the short-wave modulation by long waves
\citep{longuet1960changes,phillips1981dispersion,longuet1987propagation,henyey1988energy,zhang1990evolution},
it was not obvious from the governing equations that the fully-nonlinear
steady solutions should exist.
\citet{peureux2021unsteady} examined this problem numerically and found that
the solutions are stable only if the short-wave field is initialised from
a prior steady solution of the form of that from \citet{longuet1960changes}.
If the short-wave field is instead initialised from a uniform field, they
found that the short-wave amplitude, and thus steepness as well, grow
unsteadily.
However, here we demonstrate that the unsteady steepening of short
waves only occurs in the case of a sudden appearance of long waves, where
the velocity fields induced by the long waves act in their full capacity
on the short waves that have not yet had time to lock-in to the long-wave
crests.
In such a hypothetical scenario, sudden forcing on short waves causes the
modulation pattern to detach itself from the long-wave crests and to propagate
at the short-wave group speed.
When this occurs, the modulated short-waves are no longer in a stabilising
resonance with the surface convergence and divergence induced by the long-wave
orbital velocity, and each long-wave passage further destabilises the short-wave
field, causing it to steepen indefinitely.

Another important consideration of the hydrodynamic modulation process is
whether it should somehow be accounted for in phase-averaged spectral wave
models \citep{group1988wam,tolman1991third,donelan2012modeling} used for global
and regional wave forecasting.
Currently, most spectral wave models used for research or forecasting
do not consider the preferential steepening of short waves in the context
of their dispersion properties or the wind input
(\textit{i.e.} via the short-wave phase speed).
Some dissipation source functions use the cumulative mean squared slope
to quantify preferential dissipation of short waves due to this effect
\citep{donelan2012modeling,romero2019distribution}.
In the early days of spectral wave modelling, \citet{willebrand1975energy}
examined the effect of a wave spectrum on the short-wave group velocity,
and found the effect of up to several percent modulation.
For a practical implementation, this would incur an additional loop over all
frequencies longer than the considered short-wave frequency.
It was thus not deemed practical to include this effect in wave models at the time.
However, with the recent advances in computing power, machine learning for
acceleration of complex functions, and phase-resolved modelling of short waves,
considering the effect of wave-modulation beyond the simple cumulative mean
squared slope deserves re-examination.

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear wave theory and derive higher-order
steady solutions for the change of short-wave wavenumber, action density, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
We further examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs only for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The combined analytical and numerical framework presented here allows
quantifying the importance of various terms in the full wave conservation
equations.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx 0.2$.
As a corollary, except for a subset of initial conditions, the solutions to
the fully nonlinear crest-action conservation equations are mostly steady in
the reference frame of the long waves.
\end{abstract}

@@

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to (\ref{eq:wavenumber}), except for the last term
which represents the inhomogeneity of the group speed.
As \citet{peureux2021unsteady} explained, the presence of this term causes
unsteady growth of wave action in infinite long-wave trains.
(\ref{eq:wavenumber}) and (\ref{eq:wave_action2}) are the governing equations
that we approximate and solve analytically in \S\ref{section:analytical_solutions},
and numerically in their full form in \S\ref{section:numerical_solutions}.
It is worth noting that this system of equations is semi-coupled, meaning that
it is possible to solve (\ref{eq:wavenumber}) on its own, but solving
(\ref{eq:wave_action2}) requires also the solution of (\ref{eq:wavenumber}).
In other words, the wave kinematics (the wavenumber distribution) are not
concerned with the wave action.
In contrast, the wave action is governed, among others, by the
group velocity and thus depends on the wavenumber.

@@

\begin{equation}
\label{eq:gravity_modulation_linear_no_curvature_mean_level}
\frac{\widetilde{g}}{g} = \left( 1 - \varepsilon_L \cos{\psi} \right)
\end{equation}
which is the form used by \citet{peureux2021unsteady}, for example.

@@

\subsection{Unsteady growth in Pereux et al. (2021)}
\label{subsection:unsteady_growth}

@@

Why is it, then, that the solution for short-wave modulation is not only
quantitatively, but also qualitatively, different depending on the short-wave
initial conditions?
It may at first seem that there is something special about the uniform initial
conditions that makes its solutions grow unsteadily and indefinitely.
\citet{peureux2021unsteady} described this scenario as "...the sudden appearance
of a long wave perturbation in the middle of a homogeneous short wave field".
However, it is not obvious whether such a scenario is realistic as waves tend to
disperse into groups, and the time scales of incoming swell fronts are much
longer than those of short-wave generation by wind.
To test whether the sudden onset of long-wave forcing causes the unsteady growth,
we repeat the numerical simulation of Fig.
\ref{fig:modulation_3panel_infinite}, except for a linear ramp applied to $a_L$
(and consequently, the other long-wave properties including the orbital
velocities and the local steepness) for the first five periods:
$a_L(t) = min(a_L, a_L t / \left(5 T_L\right))$, where $T_L$ is the long-wave period.
This result is shown in Fig. \ref{fig:modulation_3panel_ramp}.
The key distinction that arises with the introduction of the ramp is that the
modulation pattern of short waves is now locked-in to the long-wave crests,
and their indefinite steepening vanishes.
The stabilisation effect of the long-wave ramp is more readily assessed in
Fig. \ref{fig:unsteady_growth_timeseries}, which shows the maximum short-wave
steepness modulation as function of time over 30 long-wave periods, for the
infinite long-wave train case and the case of applying a linear ramp.
Comparing the wave action propagation ($C_g \frac{\partial N}{\partial x}$) and
inhomogeneity ($N \frac{\partial C_g}{\partial x}$) tendencies reveals that
after only one long-wave period, the propagation and inhomogeneity tendencies
are unbalanced and amplify the steepening of short waves
(Fig. \ref{fig:inhomogeneity_tendencies}).
\citet{peureux2021unsteady} correctly pointed out that the unsteady growth of
short-wave steepness occurs due to the resonance of the inhomogeneity
tendency with the short-wave action perturbation.
However, in the case of a gentle ramp-up of the long-wave forcing, the
propagation and inhomogeneity tendencies mostly balance each other and allow
the short-wave steepening to lock-in to the long-wave crests.

@@

We now proceed to examine the numerical solutions with the long-wave forcing
applied as a long-wave group.
The wave group is simulated as a non-dispersive train of long waves whose
amplitude scales with $\sin[\pi t / (n T_L)]$, where $n$ is the number of
long waves in the group.
The elevation of the long waves thus gradually ramps up from rest to $a_L$ at
$t = n T_L / 2$, after which it gradually decreases toward zero.
In this simulation, $n = 10$ and $\varepsilon_L = 0.1$ as in the previous two
simulations.
The modulations of short-wave wavenumber, amplitude, and steepness in response
to such long-wave group is shown in Fig. \ref{fig:modulation_3panel_groups}.
Like in the case of a linear ramp, the modulation pattern here is locked in to
the long-wave crest and peaks at approximately 1.2 for short-wave steepness.
After the peak of the long-wave group passes, the short-wave modulation recedes
back toward the resting state.

@@

The solutions presented here, analytical and numerical alike, are valid only
for certain scales of short and long waves.
Specifically, the wave action conservation equation, introduced by
\citet{bretherton1968wavetrains}, requires that the ambient velocity is slowly
varying on the time scales of short waves, \textit{i.e.}
$|\partial U / \partial x| \ll k |U|$
and $|\partial U / \partial t| \ll \sigma |U|$.
This implies $k_L \ll k$ and $\sigma_L \ll \sigma$, respectively.
Another requirement is for the short-waves to remain linear, which implies
$a k \ll 1$.
Furthermore, the short waves examined here do not include the effects of
surface tension (gravity-capillary waves) or intermediate or shallow water,
although the solutions can be extended to those conditions, as
\citet{phillips1981dispersion} did, for example.
Within these limits, the crest-action conservation equations provide an
elegant and relatively simple framework to study the hydrodynamic modulation
of short waves by longer waves.
This approach, of course, isolates this particular process from other
dominant modulation and nonlinear transfer processes, such as wave growth
by wind, wave dissipation, and aerodynamic sheltering by the longer waves,
that are otherwise present in most measurements
\citep{plant1986two,laxague2017gravity}.

@@

Despite several studies that provided qualitatively similar steady solutions
for the short-wave modulation by long waves
\citep{longuet1960changes,phillips1981dispersion,longuet1987propagation,henyey1988energy,zhang1990evolution},
it was not obvious from the governing equations that the fully-nonlinear
steady solutions should exist.
\citet{peureux2021unsteady} examined this problem numerically and found that
the solutions are stable only if the short-wave field is initialised from
a prior steady solution of the form of that from \citet{longuet1960changes}.
If the short-wave field is instead initialised from a uniform field, they
found that the short-wave amplitude, and thus steepness as well, grow
unsteadily.
However, here we demonstrate that the unsteady steepening of short
waves only occurs in the case of a sudden appearance of long waves, where
the velocity fields induced by the long waves act in their full capacity
on the short waves that have not yet had time to lock-in to the long-wave
crests.
In such a scenario, sudden forcing on short waves causes the modulation
pattern to detach itself from the long-wave crests and to propagate at the
short-wave group speed.
When this occurs, the modulated short-waves are no longer in a stabilising
resonance with the surface convergence and divergence induced by the long-wave
orbital velocity, and each long-wave passage further destabilises the short-wave
field, causing it to steepen indefinitely.

Another important consideration of the hydrodynamic modulation process is
whether it should somehow be accounted for in phase-averaged spectral wave
models \citep{group1988wam,tolman1991third,donelan2012modeling} used for global
and regional wave forecasting.
Currently, most spectral wave models used for research or forecasting
do not consider the preferential steepening of short waves in the context
of their dispersion properties or the wind input via the short-wave phase speed.
Some dissipation source functions use the cumulative mean squared slope
to quantify preferential dissipation of short waves due to this effect
\citep{donelan2012modeling,romero2019distribution}.
In the early days of spectral wave modelling, \citet{willebrand1975energy}
examined the effect of a wave spectrum on the short-wave group velocity
and found the modulation effect to be up to several percent.
For a practical implementation, this would incur an additional loop over all
frequencies longer than the considered short-wave frequency.
It was thus not deemed practical to include this effect in wave models at the
time.
However, with the recent advances in computing power, machine learning for
acceleration of complex functions, and phase-resolved modelling of short waves,
considering the effect of wave-modulation beyond the simple cumulative mean
squared slope deserves re-examination.
